Preprocessing

### 2017-09-06 GWAS-biobank-scale.Rmd
use FBMs

```{r}
system.time(
  rdsfile <- snp_readBed(bedfile, backingfile = "tmp_biobank")
) # 1.6h / 1h
```

@@

```{r}
NCORES <- parallel::detectCores() / 2
# Get SNPs in long-range LD regions
system.time(
  ind.excl <- snp_indLRLDR(biobank$map$chromosome,
                           biobank$map$physical.pos)
)
# No parallelization for this huge dataset
system.time(
  ind.keep <- snp_clumping(biobank$genotypes, 
                           infos.chr = biobank$map$chromosome,
                           exclude = ind.excl,
                           ncores = 1)
) # 6.3h / 5.4h
```

>>>>

```{r}
system.time(
  rdsfile <- snp_readBed(bedfile, backingfile = "backingfiles/tmp_biobank")
) # 1.6h / 1h
```

@@

```{r}
NCORES <- nb_cores()
# Get SNPs in long-range LD regions
system.time(
  ind.excl <- snp_indLRLDR(biobank$map$chromosome,
                           biobank$map$physical.pos)
)
# No parallelization for this huge dataset
system.time(
  ind.keep <- snp_clumping(biobank$genotypes, 
                           infos.chr = biobank$map$chromosome,
                           exclude = ind.excl,
                           ncores = 1)
) # 6.3h / 5.4h
```

### 2017-09-06 autoSVD.Rmd
use FBMs

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
                      fig.width = 16, fig.asp = 0.7, error = TRUE)
options(width = 85)
```

@@

celiac <- snp_attach("backingfiles/celiacQC.rds")
G <- celiac$genotypes
CHR <- celiac$map$chromosome
POS <- celiac$map$physical.pos
NCORES <- parallel::detectCores() / 2
```

@@

```{r}
oneSNP <- which.max(abs(obj.svd$v[, 4]))
(plot_scoresA <- plot(obj.svd, type = "scores", scores = 4:5, coeff = 2) +
    aes(color = as.factor(attach.BM(G)[, ind.keep[oneSNP]])) + 
    labs(title = NULL, color = "Genotype") +
    theme(legend.position = c(0.15, 0.15)) +
    guides(color = guide_legend(override.aes = list(size = 4))))
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", 
                      fig.width = 16, fig.asp = 0.7)
options(width = 85)
```

@@

celiac <- snp_attach("backingfiles/celiacQC.rds")
G <- celiac$genotypes
CHR <- celiac$map$chromosome
POS <- celiac$map$physical.pos
NCORES <- nb_cores()
```

@@

```{r}
oneSNP <- which.max(abs(obj.svd$v[, 4]))
(plot_scoresA <- plot(obj.svd, type = "scores", scores = 4:5, coeff = 2) +
    aes(color = as.factor(G[, ind.keep[oneSNP]])) + 
    labs(title = NULL, color = "Genotype") +
    theme(legend.position = c(0.15, 0.15)) +
    guides(color = guide_legend(override.aes = list(size = 4))))
```

### 2017-09-06 benchmark-pca.Rmd
use FBMs

NCORES <- parallel::detectCores() / 2

@@

```{r}
# Write new bed files keeping only a subset of SNPs and replicating individuals
snp_writeBed(celiac, "backingfiles/celiacQC_keep1.bed",
             ind.col = ind.keep)
snp_readBed("backingfiles/celiacQC_keep1.bed", "celiacQC_keep1")

snp_writeBed(celiac, "backingfiles/celiacQC_keep5.bed",
             ind.row = rep(rows_along(G), each = 5),
             ind.col = ind.keep)
snp_readBed("backingfiles/celiacQC_keep5.bed", "celiacQC_keep5")

snp_writeBed(celiac, "backingfiles/celiacQC_keep10.bed",
             ind.row = rep(rows_along(G), each = 10),
             ind.col = ind.keep)
snp_readBed("backingfiles/celiacQC_keep10.bed", "celiacQC_keep10")
```

>>>>

NCORES <- nb_cores()

@@

```{r}
# Write new bed files keeping only a subset of SNPs and replicating individuals
snp_writeBed(celiac, "backingfiles/celiacQC_keep1.bed",
             ind.col = ind.keep)
snp_readBed("backingfiles/celiacQC_keep1.bed", "backingfiles/celiacQC_keep1")

snp_writeBed(celiac, "backingfiles/celiacQC_keep5.bed",
             ind.row = rep(rows_along(G), each = 5),
             ind.col = ind.keep)
snp_readBed("backingfiles/celiacQC_keep5.bed", "backingfiles/celiacQC_keep5")

snp_writeBed(celiac, "backingfiles/celiacQC_keep10.bed",
             ind.row = rep(rows_along(G), each = 10),
             ind.col = ind.keep)
snp_readBed("backingfiles/celiacQC_keep10.bed", "backingfiles/celiacQC_keep10")
```

### 2017-09-06 compare-GWAS-PRS-desktop.Rmd
compare desktop OK
use FBMs

Here, we suppose that the data have been pre-processed (quality control and imputation), and is available as a PLINK binary file containing 15,250 individuals and 281,122 SNPs.

```{r}
suppressMessages({
  # devtools::install_github("privefl/bigsnpr")
  library(bigsnpr)   
  # install.packages("glue")
  library(glue)      
  # install.packages("ggplot2")
  library(ggplot2)
})
# FastPCA: https://data.broadinstitute.org/alkesgroup/EIGENSOFT/EIG-6.1.4.tar.gz
smartpca <- "../EIG-6.1.4/bin/smartpca"

@@

NCORES <- parallel::detectCores() / 2
plink <- "../plink_linux_x86_64/plink"
```

@@

gwas2 <- fread2("plink.assoc.logistic")
all.equal(gwas2$P, 10^predict(gwas)) 
```

@@

# PRS
(thrs <- c(0, exp(seq(log(0.2), log(550), length.out = 50))))
lpS <- -predict(gwas.train)
system.time(
  prs <- snp_PRS(G, betas = gwas.train$estim,
                 ind.test = ind.test,
                 ind.keep = ind.keep,
                 lpS = lpS, 
                 thr.list = thrs)
)

>>>>

Here, we suppose that the data have been pre-processed (quality control and imputation), and is available as a PLINK binary file.

```{r}
suppressMessages({
  # devtools::install_github("privefl/bigsnpr")
  library(bigsnpr)   
  # install.packages("glue")
  library(glue)      
  # install.packages("ggplot2")
  library(ggplot2)
})
# FastPCA: https://data.broadinstitute.org/alkesgroup/EIGENSOFT/EIG-6.1.4.tar.gz
smartpca <- "../EIG-6.1.4/bin/smartpca"
plink <- download_plink()

@@

NCORES <- nb_cores()
```

@@

gwas2 <- fread2("plink.assoc.logistic")
all.equal(gwas2$P, predict(gwas, log10 = FALSE)) 
```

@@

# PRS
(thrs <- c(0, exp(seq(log(0.2), log(550), length.out = 50))))
lpS <- -predict(gwas.train)
system.time(
  prs <- snp_PRS(G, betas.keep = gwas.train$estim[ind.keep],
                 ind.test = ind.test,
                 ind.keep = ind.keep,
                 lpS.keep = lpS[ind.keep], 
                 thr.list = thrs)
)

### 2017-09-06 imputation.Rmd
use FBMs

NCORES <- parallel::detectCores() / 2
PLINK <- "../plink_linux_x86_64/plink"
BEAGLE <- "../../beagle.21Jan17.6cc.jar"
```

@@

# Fill a copy of the matrix with NAs (coded as 03)
X <- attach.BM(G)
XNA <- as.BM.code(deepcopy(X), code = bigsnpr:::CODE_IMPUTE_PRED)
XNA[indNA] <- as.raw(3)

GNA <- describe(XNA)
```

@@

```{r}
# Cleaning from previous runs 
file.remove(list.files("backingfiles", pattern = "tmp*", full.names = TRUE))
```

```{r}
# Write a new bedfile
popresNA <- popres
popresNA$genotypes <- GNA
bedfile <- snp_writeBed(popresNA, bedfile = "backingfiles/tmp.bed")
```

```{r}
system.time(
  bedfile2 <- snp_beagleImpute(beagle.path = BEAGLE,
                               plink.path = PLINK,
                               bedfile.in = bedfile,
                               memory.max = 30,
                               ncores = NCORES)
) # 14.6 / 14.8 / 14.8 hours
```

```{r}
# Read bedfile into R and assess the errors of imputation
popres.beagle <- snp_readBed(bedfile2, backingfile = "tmp_impute") %>%
  snp_attach()
X_beagle <- attach.BM(popres.beagle$genotypes)

mean(X_beagle[indNA] != X[indNA]) # 3.1 / 3.1 / 3.1 %
table(X_beagle[indNA], X[indNA])
```

@@

```{r, fig.asp=1}
# Errors
mean(XNA[indNA] != X[indNA]) # 4.7 / 4.8 / 4.8 %
table(XNA[indNA], X[indNA])

# Plot the estimation of the errors of imputation
data_frame(col = indNA[, 2], error = XNA[indNA] != X[indNA]) %>%
  group_by(col) %>%
  summarise(nb_err = sum(error)) %>%
  mutate(nb_err_est = (infos$pError * infos$pNA * n)[col]) %>%
  ggplot() %>%
  bigstatsr:::MY_THEME(coeff = 2) + 
  geom_abline(slope = 1, intercept = 0, col = "red") + 
  geom_point(aes(nb_err, nb_err_est), alpha = 0.4) +
  labs(x = "Number of errors", y = "Estimated number of errors")

>>>>

NCORES <- nb_cores()
```

@@

# Fill a copy of the matrix with NAs (coded as 03)
GNA <- big_copy(G)
GNA[indNA] <- as.raw(3)
```

@@

```{r}
# Write a new bedfile
popresNA <- popres
popresNA$genotypes <- GNA
bedfile <- snp_writeBed(popresNA, bedfile = tempfile(fileext = ".bed"))
```

```{r}
system.time(
  bedfile2 <- snp_beagleImpute(bedfile.in = bedfile,
                               memory.max = 30,
                               ncores = NCORES)
) # 14.6 / 14.8 / 14.8 hours
```

```{r}
# Read bedfile into R and assess the errors of imputation
popres.beagle <- snp_readBed(bedfile2, backingfile = tempfile()) %>%
  snp_attach()
G.beagle <- popres.beagle$genotypes

mean(G.beagle[indNA] != G[indNA]) # 3.1 / 3.1 / 3.1 %
table(G.beagle[indNA], G[indNA])
```

@@

```{r, fig.asp=1}
# Errors
GNA$code256 <- bigsnpr:::CODE_IMPUTE_PRED
mean(GNA[indNA] != G[indNA]) # 4.7 / 4.8 / 4.8 %
table(GNA[indNA], G[indNA])

# Plot the estimation of the errors of imputation
data_frame(col = indNA[, 2], error = GNA[indNA] != G[indNA]) %>%
  group_by(col) %>%
  summarise(nb_err = sum(error)) %>%
  mutate(nb_err_est = (infos$pError * infos$pNA * n)[col]) %>%
  ggplot() %>%
  bigstatsr:::MY_THEME(coeff = 2) + 
  geom_abline(slope = 1, intercept = 0, col = "red") + 
  geom_point(aes(nb_err, nb_err_est), alpha = 0.4) +
  labs(x = "Number of errors", y = "Estimated number of errors")

### 2017-09-06 preprocessing.Rmd
use FBMs

NCORES <- parallel::detectCores() / 2
plink <- "../plink_linux_x86_64/plink"

popresQC.bed <- snp_plinkQC(
  plink.path = plink,
  prefix.in = "../POPRES_data/POPRES_allchr",
  file.type = "--file", # ped/map
  geno = 0.05,
  mind = 0.05,
  maf = 0.05,
  hwe = 1e-10,
  autosome.only = TRUE
)

popresQC2.bed <- snp_plinkIBDQC(plink, popresQC.bed, ncores = NCORES)
```

```{r}
celiacQC.bed <- snp_plinkQC(
  plink.path = plink,
  prefix.in = "../thesis-celiac/Dubois2010_data/FinnuncorrNLITUK1UK3hap300",
  geno = 0.05,
  mind = 0.05,
  maf = 0.05,
  hwe = 1e-10,
  autosome.only = TRUE
)

celiacQC2.bed <- snp_plinkIBDQC(plink, celiacQC.bed, ncores = NCORES)
```

## Transform PLINK files in big.matrix format

```{r}
popresQC.rds <- snp_readBed(popresQC2.bed, "popresQC")

system.time(
  celiacQC.rds <- snp_readBed(celiacQC2.bed, "celiacQC")
)
```

@@

```{r}
{{G@code <- bigsnpr:::CODE_IMPUTE_PRED}}
big_counts(G, ind.col = 1:10)
```

>>>>

NCORES <- nb_cores()

popresQC.bed <- snp_plinkQC(
  prefix.in = "../POPRES_data/POPRES_allchr",
  file.type = "--file", # ped/map
  geno = 0.05,
  mind = 0.05,
  maf = 0.05,
  hwe = 1e-10,
  autosome.only = TRUE
)

popresQC2.bed <- snp_plinkIBDQC(popresQC.bed, ncores = NCORES)
```

```{r}
celiacQC.bed <- snp_plinkQC(
  prefix.in = "../thesis-celiac/Dubois2010_data/FinnuncorrNLITUK1UK3hap300",
  geno = 0.05,
  mind = 0.05,
  maf = 0.05,
  hwe = 1e-10,
  autosome.only = TRUE
)

celiacQC2.bed <- snp_plinkIBDQC(celiacQC.bed, ncores = NCORES)
```

## Transform PLINK files in bigSNP format

```{r}
popresQC.rds <- snp_readBed(popresQC2.bed, "backingfiles/popresQC")

system.time(
  celiacQC.rds <- snp_readBed(celiacQC2.bed, "backingfiles/celiacQC")
)
```

@@

```{r}
# You need to change the code of G
G$code256 <- bigsnpr:::CODE_IMPUTE_PRED
big_counts(G, ind.col = 1:10)
```

```{r}
# To make this permanent, you need to save (modify) the file on disk
celiac$genotypes$code256 <- bigsnpr:::CODE_IMPUTE_PRED
saveRDS(celiac, "backingfiles/celiacQC.rds")
```

```{r, include=FALSE}
library(gmailr)
send_message(
  mime(To = "florian.prive.21@gmail.com",
       From = "florian.prive.21@gmail.com",
       Subject = sprintf("Knitting of %s has finished",
                         knitr::current_input()))
)
```
